---
title: "Appendix Chunk"
format: pdf
---



```{r, echo = FALSE}
library(tidyverse)
library(stargazer)
library(car)

#### DATA IMPORT/WRANGLING 

# reading the csv:
RegentsExams <- read.csv("regents_scores.csv")
SchoolQualityReport <- read.csv("2017_2018_SchoolQualityReport.csv")


# selecting the relevant variables from this data set.
SchoolQuality_Selected <- SchoolQualityReport |> 
  select(DBN, School.Name, Percent.of.teachers.with.3.or.more.years.of.experience, Percent.of.Students.Chronically.Absent, Rigorous.Instruction...Percent.Positive, Supportive.Environment.Rating)

# changing one column name to make it easier to left join later
colnames(SchoolQuality_Selected)[1] <- c("School.DBN")

# filtering the RegentsExams data set 
RegentsEdited <- RegentsExams |> 
  select(2:10) |>
  filter(School.Level == "High school", Year == "2018", School.Type == "General Academic")

# filtering for only the two tests we are interested in
RegentsEdited <- RegentsEdited |> 
  filter(Regents.Exam == "Common Core Algebra2" | Regents.Exam == "Common Core English")


# creating two new rows, one for the english scores one for the math scores
RegentsEdited <- RegentsEdited |> 
  pivot_wider(names_from = Regents.Exam,
              values_from = Mean.Score)

# changing the names to work for programming: 
colnames(RegentsEdited)[8:9] <- c("Mean.Algebra2", "Mean.English")


# creating a data frame that only contains Algebra2 observations
Algebra2 <- RegentsEdited |> 
  select(!9) |>
  filter(Mean.Algebra2 != "s" & !is.na(Mean.Algebra2))

English <- RegentsEdited |> 
  select(School.DBN, Mean.English) |>
  filter(Mean.English != "s" & !is.na(Mean.English))

# joining the math and english exam data sets into final one 
FinalRegents <- left_join(x = Algebra2, y = English, by = "School.DBN")

# now I have a working data frame for the exam scores. next step is to join the two data sets together to make the final one for analysis


# selecting only the minimum columns we are interested in
Scores <- FinalRegents |> 
  select(School.DBN, Mean.Algebra2, Mean.English)

# left joining by School.DBN to get the ultimate real final data frame we want to work with: 
FinalData <- left_join(x = Scores, y = SchoolQuality_Selected, by = "School.DBN")

# The test score columns were character vectors so we needed to coerce them to numeric vectors.
FinalData$Mean.Algebra2 <- as.numeric(FinalData$Mean.Algebra2)
FinalData$Mean.English <- as.numeric(FinalData$Mean.English)

# need to re-code the column names to make them easier to code with
names <- names(FinalData)
snake <- snakecase::to_snake_case(names)
colnames(FinalData) <- snake
renamed <- c("school_dbn", "mean_algebra2", "mean_english", "school_name", "teacher_experience", "chronically_absent", "rigorous_instruction", "supportive_environment")
colnames(FinalData) <- renamed

# final data set is called FinalData

#### CONDITION CHECKS/INITIAL INVESTIGATION FOR MODEL 1: 

## investigating the english model: 
# fitting the model: 
lm_english <- lm(mean_english ~ teacher_experience + chronically_absent + rigorous_instruction + supportive_environment, data = FinalData)

summary(lm_english)
# for model 1 we noticed that only the intercept and the variable chronically_absent are statistically significant.

# condition checks: 
plot(lm_english, which = 1)
plot(lm_english, which = 2)
vif(lm_english)
# we noticed that these residuals are not normally distributed 

#### LOOKING AT INDIVIDUAL VARIABLES (MODEL 1):
ggplot(data = FinalData, aes(x = chronically_absent, y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Chronically Absent vs. Mean English")
ggplot(data = FinalData, aes(x = teacher_experience, y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Teacher Experience vs. Mean English")
ggplot(data = FinalData, aes(x = rigorous_instruction, y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Rigorous Instruction vs. Mean English")
# chronically absent vs mean english was the only scatter plot to look remotely linear. 

ggplot(data = FinalData, aes(x = log(chronically_absent), y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Log Chronically Absent vs. Mean English")
ggplot(data = FinalData, aes(x = log(teacher_experience), y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Log Teacher Experience vs. Mean English")
ggplot(data = FinalData, aes(x = log(rigorous_instruction), y = mean_english)) + geom_point() + geom_smooth() +
  ggtitle("Log Rigorous Instruction vs. Mean English")
# noticed that applying log to the variables didn't fix anything, so we kept variables unchanged. 


#### CONDITION CHECKS/INITIAL INVESTIGATION FOR MODEL 2: 
## investigating the math model: 
# fitting the model: 
lm_math <- lm(mean_algebra2 ~ teacher_experience + chronically_absent + rigorous_instruction + supportive_environment, data = FinalData)

summary(lm_math)
# for model 2 we noticed that the intercept, teacher_experience, and chronically_absent were statistically significant. 

# condition checks: 
plot(lm_math, which = 1)
plot(lm_math, which = 2)
vif(lm_math)
# we noticed that these residuals are not normally distributed 

#### LOOKING AT INDIVIDUAL VARIABLES (MODEL 2):
ggplot(data = FinalData, aes(x = chronically_absent, y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Chronically Absent vs. Mean Math")
ggplot(data = FinalData, aes(x = teacher_experience, y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Teacher Experience vs. Mean Math")
ggplot(data = FinalData, aes(x = rigorous_instruction, y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Rigorous Instruction vs. Mean Math")
# chronically absent vs mean algebra2 was the only scatter plot to look remotely linear. 

ggplot(data = FinalData, aes(x = log(chronically_absent), y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Log Chronically Absent vs. Mean Math")
ggplot(data = FinalData, aes(x = log(teacher_experience), y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Log Teacher Experience vs. Mean Math")
ggplot(data = FinalData, aes(x = log(rigorous_instruction), y = mean_algebra2)) + geom_point() + geom_smooth() +
  ggtitle("Log Rigorous Instruction vs. Mean Math")
# noticed that applying log to the variables didn't fix anything, so we kept variables unchanged. 
```

make prediction plot for chronically absent, one for each model. USE what was stat signifiant. 

```{r}
# making predictions for english lm 
cleaned <- FinalData |>
  filter(!is.na(teacher_experience))

pred_dataenglish <- with(cleaned, 
                         data.frame(teacher_experience = mean(teacher_experience),
                                    chronically_absent = seq(from = 0, to = .75, by = 0.025),
                                    rigorous_instruction = mean(rigorous_instruction),
                                    supportive_environment = "Meeting Target"))

# generating predictions: 
predictions_eng <- as.data.frame(predict(lm_english, pred_dataenglish, type = "response", se.fit = TRUE))

# inserting the standard errors: 
pred_dataenglish$pred_prob <- predictions_eng$fit
pred_dataenglish$pred_lower <- predictions_eng$fit-(1.96*predictions_eng$se.fit)
pred_dataenglish$pred_upper <- predictions_eng$fit+(1.96*predictions_eng$se.fit)

ggplot(data = pred_dataenglish, aes(x = chronically_absent, y = pred_prob)) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), fill = "red", alpha = 0.50) +
  geom_line() + 
  labs(title = "Average English Regents Score vs. Chronic Absence Rates", 
       x = "Chronic Absence Rate", 
       y = "Average English Regents Score") +
  theme_minimal()
```

```{r}
# making predictions for math lm 
pred_datamath <- with(cleaned, 
                      data.frame(teacher_experience = mean(teacher_experience),
                                 chronically_absent = seq(from = 0, to = .75, by = 0.025),
                                 rigorous_instruction = mean(rigorous_instruction),
                                 supportive_environment = "Meeting Target"))

# generating predictions: 
predictions_mth <- as.data.frame(predict(lm_math, pred_datamath, type = "response", se.fit = TRUE))

# inserting the standard errors: 
pred_datamath$pred_prob <- predictions_mth$fit
pred_datamath$pred_lower <- predictions_mth$fit-(1.96*predictions_mth$se.fit)
pred_datamath$pred_upper <- predictions_mth$fit+(1.96*predictions_mth$se.fit)

ggplot(data = pred_datamath, aes(x = chronically_absent, y = pred_prob)) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper),fill = "blue", alpha = 0.50) +
  geom_line() + 
  labs(title = "Average Algebra 2 Regents Score vs. Chronic Absence Rates", 
       x = "Chronic Absence Rate", 
       y = "Average Algebra 2 Regents Score") +
  theme_minimal()
```

```{r, results='asis'}
stargazer(lm_math, 
          type = "latex", 
          header = F, 
          float = F)
```

```{r}
summary(lm_math)
```
**actual interpretation for math:**

From the model predicting a school's average 2018 Algebra 2 Regents score, two statistically significant variables emerge as predictors of math test scores for a school. First, as hypothesized, teachers' teaching experience is positively associated with high school students' math scores. This implies that as teachers' experience increases, students' math scores also increase. However, the school's supportive environment, the other variable mentioned in our initial hypothesis, did not exhibit statistical significance in this model. We failed to provide evidence for an association between the school's supportive environment and students' math scores. Nevertheless, the coefficient for the control variable "chronic absence" is statistically significant. There is a negative association between chronic absence and students' math scores, indicating that as students' absences increase, their math scores decrease.

The adjusted $R^2$ for this model is 46.5%, signifying that the model explains a moderate proportion of the variability in students' math scores, leaving room for improvement. The model's F-statistic is also statistically significant, indicating that it successfully reduces the unexplained variation in students' math scores by incorporating four different variables.

While this model validates half of our initial hypothesis by confirming a significant association between math test scores and teachers' teaching experiences, it falls short in supporting a positive relationship between math scores and schools' supportive environment.


?????

```{r, results='asis'}
stargazer(lm_math, lm_english,
          type = "latex", 
          header = F, 
          float = F)
```

```{r}
summary(lm_english)
```
**actual interpretation for english:**

For our model predicting a school's average 2018 English Regents score only the coefficient estimate for chronic absenteeism was statistically significant and negatively associated with the dependent variable. As the percentage of chronic absenteeism within a student body increases, a school's average score for the 2018 English Regents decreases. The other predictors included in this model were not statistically significant so our initial hypothesis that average English test scores have a relationship with teacher experience and a school's supportive environment was not supported by this model.

Although our model did not support our hypothesis, it was moderately successful at explaining the variation in our dependent variable, producing an adjusted $R^2$ of $55.75$%. Additionally this model's F-stat was statistically significant meaning this model has explanatory power and managed to reduce the unexplained variability of our dependent variable. 


# Conclusion
For this study, we would like to see whether a school's average Algebra 2 and English Regents scores are significantly associated with a school's supportive environment, teachers' teaching experience, the percentage of rigorous instruction, and students' absenteeism rates. 

Our analysis does not supports our primary hypothesis. In our initial model, where the dependent variable is a school's average English Regents score, we failed to reject the null hypothesis for all independent variables except the factor of chronic absenteeism. This indicates there is a significant relationship between the average English Regents score in schools and the percentage of students experiencing chronic absenteeism.

Moving on to the second model, with the schools' average Algebra 2 Regents score as the dependent variable, we failed to reject the null hypothesis for both the. teacher experience and supportive environment predictors. However, we reject the null hypothesis for rigorous instruction and chronic absenteeism, indicating that there is a relationship between a school's average Algebra 2 Regents score and chronic absenteeism and rigorous teaching. We can successfully conclude that the chronically absent factor is statistically significant across both models.

# Limitation:
In our research, we selected only four variables as predictor variables, omitting consideration of other potential factors such as school race composition and overall school quality that could have an impact on our results. This selective approach may have led to an oversimplification of our sample, as we focused solely on students' performance measured by a school's average Algebra 2 and English Regents scores. 

Furthermore, it's essential to note that we combined two distinct data sets, since our first data set only included students' exam scores from before entering high school. This merging of data sets introduces potential complexities and nuances that should be acknowledged in interpreting our findings.

Since we also failed to pass the initial regression conditions we need to be very cautious about making any causal inferences based on our results since they are not technically valid. This problem can be addressed in the future by selecting data with more observations, and be more careful when selecting predictors. These adjustments can enhance the quality of our models and increase the accuracy of our findings.
